```{r}
## Use install.packages() to install these packages:
suppressMessages(library(tidyverse))
suppressMessages(library(janitor))
suppressMessages(library(plotly))
suppressMessages(library(ggrepel))
suppressMessages(library(purrr))
suppressMessages(library(dplyr))

## Use BiocManager::install() to install these packages:
suppressMessages(library(DESeq2))
suppressMessages(library(GeneTonic))
suppressMessages(library(pcaExplorer))
suppressMessages(library(topGO))
suppressMessages(library(tidybulk))
suppressMessages(library(AnnotationDbi))
suppressMessages(library(org.Hs.eg.db))
suppressMessages(library(SummarizedExperiment))
suppressMessages(library(dittoSeq))
suppressMessages(library(msigdbr))
suppressMessages(library(fgsea))

library(tidybulk)
library(ggplot2)
library(ggrepel)
library(SummarizedExperiment)
library(dplyr)
```

```{r}
# Load required libraries
library(tidyverse)

# Define base directory containing sample folders
base_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/rnaseq/Gene_expresion_analysis/quant_sf_gene_exp_level_files"

# Get full paths to sample directories
sample_dirs <- list.dirs(base_dir, recursive = FALSE, full.names = TRUE)

# Extract sample folder names
sample_names <- basename(sample_dirs)

# Construct metadata dataframe
sample_metadata <- tibble(
  sample = sample_names,
  path = file.path(sample_dirs, "quant.sf")
) %>%
  # Derive structured metadata
  mutate(
    cell_line = case_when(
      str_detect(sample, "^NALM6") ~ "NALM6",
      str_detect(sample, "^REH") ~ "REH",
      str_detect(sample, "^Jurkat") ~ "Jurkat",
      str_detect(sample, "^MOLM13") ~ "MOLM13",
      TRUE ~ NA_character_
    ),
    genotype = case_when(
      str_detect(sample, "TT") ~ "TT",
      str_detect(sample, "TC") ~ "TC",
      str_detect(sample, "CC") ~ "CC",
      TRUE ~ "WT"
    ),
    clone = str_extract(sample, "Clone[0-9]+") %>% str_remove("Clone") %>% as.integer(),
    replicate = 1,
    # Combine into full sample ID (e.g. REH_TT_1)
    sample_id_full = paste(cell_line, genotype, ifelse(is.na(clone), "NA", clone), sep = "_")
  )

# Set rownames to match combined sample ID
rownames(sample_metadata) <- sample_metadata$sample_id_full
```

# 🧬 2. Load the gene annotation (GTF → tx2gene)

```{r}
library(rtracklayer)
library(tidyverse)

# Load GTF annotation
gtf <- import("/Users/cojulian/Desktop/Tzu_projects/Matt/reference_genomces/Homo_sapiens.GRCh38.113.gtf.gz")
gtf_df <- as.data.frame(gtf)

# Create transcript-to-gene mapping
tx2gene_df <- gtf_df %>%
  dplyr::select(transcript_id, gene_id, gene_name) %>%
  dplyr::distinct(transcript_id, gene_id, gene_name) %>%
  drop_na()

```

# 📁 3. Use your existing metadata (sample_metadata)

```{r}
# Assume sample_metadata already exists with these columns:
# sample, path, cell_line, genotype, clone, replicate, sample_id_full

# For compatibility with next steps
human_coldata <- sample_metadata %>%
  mutate(files = path) %>%
  select(sample_id_full, files, everything()) %>%
  rename(Sample = sample_id_full)

# Set Sample as rownames (optional)
rownames(human_coldata) <- human_coldata$Sample

```

```{r}
# Read first quant.sf file and map transcripts to genes
tmp <- read.delim(human_coldata$files[1]) %>%
  inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
  group_by(gene_id) %>%
  summarise(NumReads = as.integer(sum(NumReads)))

# Prepare matrix dimensions
num.row <- nrow(tmp)
num.col <- nrow(human_coldata)

# Initialize empty gene-level count matrix
sf.df <- data.frame(matrix(0, nrow = num.row, ncol = num.col))
colnames(sf.df) <- human_coldata$Sample
rownames(sf.df) <- tmp$gene_id

```

```{r}
for (i in seq_along(human_coldata$files)) {
  cat("Reading:", human_coldata$files[i], "\n")
  
  sf.df[, i] <- read.delim(human_coldata$files[i]) %>%
    inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
    group_by(gene_id) %>%
    summarise(NumReads = as.integer(sum(NumReads))) %>%
    pull(NumReads)
}
```

```{r}
# Convert to matrix for downstream use (DESeq2, edgeR, etc.)
Salmon.matrix <- as.matrix(sf.df)

# Confirm sample IDs match
all(colnames(Salmon.matrix) %in% rownames(human_coldata))  # Should return TRUE
```

```{r}
# 🧮 Set filtering thresholds
expression_threshold <- 10
min_clones_expressed <- 2

# 📍 Convert count matrix to long format
Salmon_long <- as.data.frame(Salmon.matrix) %>%
  tibble::rownames_to_column("gene_id") %>%
  tidyr::pivot_longer(-gene_id, names_to = "Sample", values_to = "expression")

# 🔗 Join with metadata using Sample column (no need for rownames)
Salmon_long <- Salmon_long %>%
  dplyr::left_join(
    human_coldata %>%
      dplyr::select(Sample, cell_line, genotype, clone, replicate),
    by = "Sample"
  ) %>%
  dplyr::mutate(group = paste(cell_line, genotype, sep = "_"))

# 🧮 Count how many clones per group express each gene > threshold
gene_group_counts <- Salmon_long %>%
  dplyr::group_by(gene_id, group) %>%
  dplyr::summarise(n_expressed_clones = sum(expression > expression_threshold), .groups = "drop")

# ✅ Count number of clones per group
group_sizes <- human_coldata %>%
  dplyr::mutate(group = paste(cell_line, genotype, sep = "_")) %>%
  dplyr::count(group, name = "n_clones")

# ✅ Final filter: keep genes expressed in ≥ min_clones_expressed in at least one group
genes_to_keep <- gene_group_counts %>%
  dplyr::left_join(group_sizes, by = "group") %>%
  dplyr::filter(n_expressed_clones >= pmin(n_clones, min_clones_expressed)) %>%
  dplyr::distinct(gene_id)

# ✂️ Filter original matrix
sf.df.data <- Salmon.matrix %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene_id") %>%
  dplyr::filter(gene_id %in% genes_to_keep$gene_id) %>%
  tibble::column_to_rownames("gene_id")

# ✅ Inspect result
cat("\nFinal filtered matrix dimensions:\n")
print(dim(sf.df.data))

cat("\nPreview filtered matrix:\n")
print(head(sf.df.data[, 1:5]))



```

# QC of before and after

```{r}
# 📊 1. Compare expression distribution before vs after filtering

# Unfiltered matrix to long format
Salmon_long_dist_unfiltered <- as.data.frame(Salmon.matrix) %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "expression") %>%
  mutate(stage = "Unfiltered")

# Filtered matrix to long format
Salmon_long_dist_filtered <- sf.df.data %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "expression") %>%
  mutate(stage = "Filtered")

# Combine and log-transform expression
combined_expression <- bind_rows(Salmon_long_dist_unfiltered, Salmon_long_dist_filtered) %>%
  mutate(log_expression = log2(expression + 1))

# Density plot
ggplot(combined_expression, aes(x = log_expression, fill = stage)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Log2(Expression) Distribution Before vs After Filtering",
    x = "log2(Expression + 1)",
    y = "Density",
    fill = "Stage"
  ) +
  theme_minimal()

# Boxplot
ggplot(combined_expression, aes(x = stage, y = log_expression, fill = stage)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  labs(
    title = "Boxplot of Log2(Expression) Before vs After Filtering",
    x = "",
    y = "log2(Expression + 1)"
  ) +
  theme_minimal()

# Summary
cat("Number of genes before filtering:", nrow(Salmon.matrix), "\n")
cat("Number of genes after filtering:", nrow(sf.df.data), "\n")
cat("Percentage retained:", round(nrow(sf.df.data) / nrow(Salmon.matrix) * 100, 2), "%\n")

```

# Save FPKM
```{r}
# 🧬 1. Extract gene lengths from GTF (sum of exon widths per gene)
gene_lengths <- gtf_df %>%
  filter(type == "exon") %>%
  group_by(gene_id) %>%
  summarise(gene_length_kb = sum(width) / 1000, .groups = "drop")  # convert to kilobases

# 🧮 2. Get library sizes (total counts per sample in millions)
lib_sizes <- colSums(sf.df.data) / 1e6  # convert to millions
# 📊 3. Compute FPKM
fpkm_matrix <- sweep(sf.df.data, 2, lib_sizes, "/") %>%  # divide by library size
  as.data.frame() %>%
  tibble::rownames_to_column("gene_id") %>%
  left_join(gene_lengths, by = "gene_id") %>%
  mutate(across(-c(gene_id, gene_length_kb), ~ .x / gene_length_kb))  # divide by gene length

# 🧾 4. Add gene symbols
fpkm_matrix_annotated <- fpkm_matrix %>%
  left_join(tx2gene_df %>% select(gene_id, gene_name) %>% distinct(gene_id, .keep_all = TRUE), by = "gene_id") %>%
  relocate(gene_name, .after = gene_id)  # move gene_name next to gene_id

# 🧹 5. Clean up and save
fpkm_matrix_clean <- fpkm_matrix_annotated %>%
  select(-gene_length_kb) %>%
  distinct(gene_name, .keep_all = TRUE) %>%  # remove duplicates if any
  column_to_rownames("gene_name")  # use gene symbols as rownames

# 💾 6. Save to CSV
write.csv(fpkm_matrix_clean, file = "fpkm_matrix_with_symbols.csv", row.names = TRUE)

```


# 🧬 PCA — using your metadata (cell_line, genotype)

```{r}
# 1. Filter metadata to match filtered expression matrix
filtered_coldata <- human_coldata[colnames(sf.df.data), ]

# 2. Create SummarizedExperiment
se <- SummarizedExperiment(
  assays = list(counts = as.matrix(sf.df.data)),
  colData = filtered_coldata
)

# 3. Tidybulk pipeline: abundance filter → normalize → PCA
pca_result <- se %>%
  tidybulk::keep_abundant() %>%
  tidybulk::scale_abundance() %>%
  tidybulk::reduce_dimensions(method = "PCA", .dims = 15)

# 4. PCA plot: color by genotype, shape by cell_line
pca_result %>%
  tidybulk::pivot_sample() %>%
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = genotype,
    shape = cell_line,
    label = Sample  # Optional: or use paste(cell_line, genotype, clone)
  )) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(max.overlaps = 5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PCA of Filtered Gene Expression Data",
    subtitle = "Colored by Genotype, Shaped by Cell Line",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )

## No MT genes
# Add gene_id as a column
sf.df.data_symbols <- sf.df.data %>%
    tibble::rownames_to_column("gene_id")

# Join with tx2gene_df to get gene_name
sf.df.data_symbols <- sf.df.data_symbols %>%
    left_join(
        tx2gene_df %>%
            select(gene_id, gene_name) %>%
            distinct(gene_id, .keep_all = TRUE),
        by = "gene_id"
    ) %>%
    relocate(gene_name, .after = gene_id)


# 1. Identify and remove mitochondrial genes
mt_genes <- sf.df.data_symbols %>%
  dplyr::filter(grepl("^MT-", gene_name)) %>%
  pull(gene_id)

sf.df.data_nomt <- sf.df.data_symbols %>%
  dplyr::filter(!gene_id %in% mt_genes)

# 2. Set gene_id as rownames
expr_nomt <- sf.df.data_nomt %>%
  tibble::column_to_rownames("gene_id") %>%
  dplyr::select(-gene_name)

# 3. Filter metadata to match expression matrix
filtered_coldata_nomt <- human_coldata[colnames(expr_nomt), ]

# 4. Create SummarizedExperiment
se_nomt <- SummarizedExperiment(
  assays = list(counts = as.matrix(expr_nomt)),
  rowData = sf.df.data_nomt %>% select(gene_id, gene_name) %>% tibble::column_to_rownames("gene_id"),
  colData = filtered_coldata_nomt
)

# 5. Tidybulk pipeline: abundance filter → normalize → PCA
pca_nomt <- se_nomt %>%
  tidybulk::keep_abundant() %>%
  tidybulk::scale_abundance() %>%
  tidybulk::reduce_dimensions(method = "PCA", .dims = 15)

# 6. PCA plot: color by genotype, shape by cell_line
pca_nomt %>%
  tidybulk::pivot_sample() %>%
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = genotype,
    shape = cell_line,
    label = Sample
  )) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(max.overlaps = 5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PCA of Gene Expression (MT Genes Removed)",
    subtitle = "Colored by Genotype, Shaped by Cell Line",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )



```

### PCA for NALM6

```{r}
library(tidyverse)
library(tidybulk)
library(ggrepel)

# 1. Filter metadata for NALM6 samples
nalm6_coldata <- human_coldata %>%
  filter(cell_line == "NALM6")

# 2. Filter expression matrix to match NALM6 samples
nalm6_expr <- sf.df.data_nomt[, c("gene_id", nalm6_coldata$Sample)]

# 3. Reshape expression matrix to tidy format
nalm6_expr_tidy <- nalm6_expr %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "Sample",
    values_to = "abundance"
  ) %>%
  rename(gene = gene_id)

# 4. Merge with metadata
nalm6_tidy <- nalm6_expr_tidy %>%
  left_join(nalm6_coldata, by = c("Sample" = "Sample"))

# 5. Tidybulk pipeline: explicitly define tidybulk columns
pca_nalm6 <- nalm6_tidy %>%
  tidybulk(
    .sample = Sample,
    .transcript = gene,
    .abundance = abundance
  ) %>%
  keep_abundant() %>%
  scale_abundance() %>%
  reduce_dimensions(method = "PCA", .dims = 8)

# 6. PCA plot: color by genotype, shape by clone
pca_nalm6 %>%
  pivot_sample() %>%
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = genotype,
    shape = as.factor(clone),
    label = Sample
  )) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(max.overlaps = 5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PCA of NALM6 Gene Expression",
    subtitle = "Colored by Genotype, Shaped by Clone",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )


# What is driving pc1?
# --- What is driving PC1? ---

# 1. Extract scaled expression matrix from tidybulk object
scaled_expr <- pca_nalm6 %>%
  select(gene, Sample, abundance_scaled) %>%
  pivot_wider(names_from = Sample, values_from = abundance_scaled) %>%
  column_to_rownames("gene")

# 2. Run PCA manually
pca_manual <- prcomp(t(scaled_expr), center = TRUE, scale. = FALSE)

# 3. Extract loadings (gene contributions to PCs)
loadings_df <- as.data.frame(pca_manual$rotation) %>%
  rownames_to_column("gene_id")

# 4. Get top genes driving PC1
top_pc1_genes <- loadings_df %>%
  arrange(desc(abs(PC1))) %>%
  slice_head(n = 20)

# 5. Annotate with gene symbols
top_pc1_genes_annotated <- top_pc1_genes %>%
  left_join(sf.df.data_symbols %>% select(gene_id, gene_name), by = "gene_id")

# 6. View results
print(top_pc1_genes_annotated)

# --- Remove mitochondrial genes ---

# 1. Identify mitochondrial genes
mt_genes <- sf.df.data_symbols %>%
  filter(str_detect(gene_name, "^MT-")) %>%
  pull(gene_id)

# 2. Filter out mitochondrial genes
sf.df.data_nomt <- sf.df.data_symbols %>%
  filter(!gene_id %in% mt_genes)

# 3. Prepare expression matrix without MT genes
expr_nomt <- sf.df.data_nomt %>%
  column_to_rownames("gene_id") %>%
  select(-gene_name)

# 4. Filter metadata to match expression matrix
filtered_coldata_nomt <- human_coldata[colnames(expr_nomt), ]

# 5. Create tidybulk object
tidy_nomt <- expr_nomt %>%
  rownames_to_column("gene") %>%
  pivot_longer(cols = -gene, names_to = "Sample", values_to = "abundance") %>%
  left_join(filtered_coldata_nomt, by = c("Sample" = "Sample")) %>%
  tidybulk(
    .sample = Sample,
    .transcript = gene,
    .abundance = abundance
  )

# 6. Tidybulk pipeline: abundance filter → normalize → PCA
pca_nomt <- tidy_nomt %>%
  keep_abundant() %>%
  scale_abundance() %>%
  reduce_dimensions(method = "PCA", .dims = 15)

# 7. PCA plot: color by genotype, shape by cell_line
pca_nomt %>%
  pivot_sample() %>%
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = genotype,
    shape = cell_line,
    label = Sample
  )) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(max.overlaps = 5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PCA of Gene Expression (MT Genes Removed)",
    subtitle = "Colored by Genotype, Shaped by Cell Line",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )



```



# ✅ Updated QC Suite 
## Build SummarizedExperiment with filtered matrix + metadata
```{r}
# Build SE object
sf.df.se <- SummarizedExperiment(
  assays = list(counts = as.matrix(sf.df.data)),
  colData = human_coldata[colnames(sf.df.data), ]  # Ensure matching sample order
)

# Create detailed sample names
sf.df.se$detailed_name <- paste(sf.df.se$cell_line, sf.df.se$genotype, sf.df.se$clone, sep = "_")
```


```{r}
# Calculate library sizes
library_sizes <- colSums(assay(sf.df.se))

# Plot
ggplot(
  data.frame(Sample = sf.df.se$detailed_name, LibrarySize = library_sizes),
  aes(x = Sample, y = LibrarySize)
) +
  geom_bar(stat = "identity", fill = "skyblue") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold")
  ) +
  labs(
    title = "Library Size Distribution",
    x = "Sample (cell_line_genotype_clone)",
    y = "Library Size (Raw Counts)"
  )

```



```{r}
# Log-transform
log_counts <- log2(assay(sf.df.se) + 1)

# Melt for ggplot
log_counts_long <- reshape2::melt(log_counts)
colnames(log_counts_long) <- c("Gene", "Sample", "LogExpression")

# Add cell_line/genotype/clone to annotate by color
log_counts_long <- log_counts_long %>%
  mutate(Group = sf.df.se$detailed_name[match(Sample, colnames(log_counts))])

# Plot
ggplot(log_counts_long, aes(x = LogExpression, color = Group)) +
  geom_density(size = 1, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Log2 Gene Expression Distribution by Sample",
    x = "Log2(Counts + 1)",
    y = "Density"
  ) +
  theme(
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9)
  )

```

# Doing DE, only one comparison

```{r}
# DESeq2 Analysis: NALM6 TT vs CC

# 1. Filter metadata to only NALM6 TT and CC clones
meta_data <- sample_metadata %>%
  dplyr::filter(cell_line == "NALM6", genotype %in% c("TT", "CC")) %>%
  dplyr::mutate(genotype = factor(genotype, levels = c("CC", "TT")))  # Reference = CC

# 2. Inspect sample IDs and count matrix columns
print("Sample IDs in metadata:")
print(meta_data$sample_id_full)

print("Column names in count matrix:")
print(colnames(sf.df.data))

# 3. Check for missing samples
missing_samples <- setdiff(meta_data$sample_id_full, colnames(sf.df.data))
if (length(missing_samples) > 0) {
  warning("Missing samples in count matrix: ", paste(missing_samples, collapse = ", "))
}

# 4. Subset count matrix to match metadata
filtered_bulk_matrix <- sf.df.data[, meta_data$sample_id_full]

# 5. Check dimensions
print(paste("Filtered matrix dimensions:", paste(dim(filtered_bulk_matrix), collapse = " x ")))
print(paste("Metadata dimensions:", paste(dim(meta_data), collapse = " x ")))

# 6. Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(filtered_bulk_matrix),
  colData = meta_data,
  design = ~ genotype
)

# 7. Run DE analysis
dds <- DESeq(dds)

# 8. Extract results: TT vs CC
res <- results(dds, contrast = c("genotype", "TT", "CC"), alpha = 0.05) 

# 9. Convert results to dataframe
de_results_tt_vs_cc <-  as.data.frame(res) %>%
    tibble::rownames_to_column("gene_id") %>%
    left_join(
        tx2gene_df %>%
            select(gene_id, gene_name) %>%
            distinct(gene_id, .keep_all = TRUE),
        by = "gene_id"
    ) %>%
    mutate(FoldChange = 2^log2FoldChange) %>%
    mutate(significance = case_when(
        FoldChange >= 1.5 & padj <= 0.05 ~ "up",
        FoldChange <= (1 / 1.5) & padj <= 0.05 ~ "down",
        TRUE ~ "ns"
    ))




# 10. Inspect result summary
summary(res)

```

```{r}
# Annotate with gene names (tx2gene_df assumed loaded)
volcano_data <- de_results_tt_vs_cc %>%
  # Join with gene name annotations
  left_join(tx2gene_df %>% select(gene_id, gene_name), by = "gene_id") %>%
  
  # Resolve gene_name conflict and create GeneSymbol
  mutate(GeneSymbol = ifelse(is.na(gene_name.y), gene_id, gene_name.y)) %>%
  
  # Remove duplicate gene entries
  distinct(gene_id, .keep_all = TRUE) %>%
  
  # Select relevant columns
  select(GeneSymbol, log2FoldChange, padj) %>%
  
  # Calculate FoldChange
  mutate(FoldChange = 2^log2FoldChange) %>%
  
  # Classify genes by significance and direction
  mutate(GeneType = case_when(
    FoldChange >= 1.5 & padj <= 0.05 ~ "up",
    FoldChange <= (1/1.5) & padj <= 0.05 ~ "down",
    TRUE ~ "ns"
  ))


# Define style
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

# Plot
volcano_plot <- ggplot(volcano_data, aes(
  x = log2(FoldChange),
  y = -log10(padj),
  fill = GeneType,
  size = GeneType,
  alpha = GeneType
)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(log2(1 / 1.5), log2(1.5)), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  ggrepel::geom_text_repel(
    data = filter(volcano_data, GeneType != "ns"),
    aes(label = GeneSymbol),
    size = 5,
    color = "darkblue",
    max.overlaps = 10,
    min.segment.length = Inf
  ) +
  labs(
    title = "Volcano Plot: NALM6 TT vs CC (padj ≤ 0.05, FC ≥ 1.5)",
    x = "log2(Fold Change)",
    y = "-log10(Adjusted P-value)"
  ) +
  theme_minimal(base_family = "Arial") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

# Show the plot
print(volcano_plot)
```

# All around NALM6

## Loop to run DEG NALM6 ONLY

```{r}
# 📂 1. Create output folder
out_dir <- "deg_results_NALM6"
if (!dir.exists(out_dir)) dir.create(out_dir)

# 🧬 2. Function to run DESeq2 for a given comparison with FPKM
run_deg <- function(count_matrix, metadata, ref_genotype, test_genotype, out_prefix) {
  
  # 📂 Filter metadata for NALM6 and selected genotypes
  meta_data <- metadata %>%
    filter(cell_line == "NALM6", genotype %in% c(ref_genotype, test_genotype)) %>%
    mutate(genotype = factor(genotype, levels = c(ref_genotype, test_genotype)))  # reference first
  
  # 📊 Subset count matrix
  filtered_counts <- count_matrix[, meta_data$sample_id_full]
  
  # 🧬 Get gene lengths from GTF (sum of exon widths per gene)
  gene_lengths <- gtf_df %>%
    filter(type == "exon") %>%
    group_by(gene_id) %>%
    summarise(gene_length_kb = sum(width) / 1000, .groups = "drop")  # kilobases
  
  # 📊 Get library sizes (total counts per sample in millions)
  lib_sizes <- colSums(filtered_counts) / 1e6
  
  # 🧮 Compute FPKM
  fpkm_matrix <- sweep(filtered_counts, 2, lib_sizes, "/") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene_id") %>%
    left_join(gene_lengths, by = "gene_id") %>%
    mutate(across(-c(gene_id, gene_length_kb), ~ .x / gene_length_kb))
  
  # 📈 Compute group-specific average FPKM
  fpkm_long <- fpkm_matrix %>%
    pivot_longer(-c(gene_id, gene_length_kb), names_to = "Sample", values_to = "fpkm") %>%
    left_join(meta_data %>% select(Sample = sample_id_full, genotype), by = "Sample")
  
  avg_fpkm_grouped <- fpkm_long %>%
    group_by(gene_id, genotype) %>%
    summarise(avg_fpkm = mean(fpkm, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = genotype, values_from = avg_fpkm, names_prefix = "fpkm_")
  
  # 🧪 DESeq2 analysis
  dds <- DESeqDataSetFromMatrix(
    countData = as.matrix(filtered_counts),
    colData = meta_data,
    design = ~ genotype
  )
  dds <- DESeq(dds)
  res <- results(dds, contrast = c("genotype", test_genotype, ref_genotype), alpha = 0.05)
  
  # 🧾 Format results
  deg_df <- as.data.frame(res) %>%
    tibble::rownames_to_column("gene_id") %>%
    left_join(tx2gene_df %>% select(gene_id, gene_name) %>% distinct(gene_id, .keep_all = TRUE), by = "gene_id") %>%
    mutate(FoldChange = 2^log2FoldChange) %>%
    mutate(significance = case_when(
      FoldChange >= 1.5 & padj <= 0.05 ~ "up",
      FoldChange <= (1 / 1.5) & padj <= 0.05 ~ "down",
      TRUE ~ "ns"
    )) %>%
    left_join(avg_fpkm_grouped, by = "gene_id")  # ⬅️ Add group-specific FPKM
  
  # 💾 Save results
  write.csv(deg_df, file = file.path(out_dir, paste0(out_prefix, ".csv")), row.names = FALSE)
  
  # Return the DEG dataframe invisibly
  invisible(deg_df)
}


# 🚀 3. Run for the 3 desired comparisons
deg_tt_vs_tc <- run_deg(sf.df.data, sample_metadata, "TC", "TT", "DEG_gen_exp_NALM6_TT_vs_TC")
deg_tc_vs_cc <- run_deg(sf.df.data, sample_metadata, "CC", "TC", "DEG_gen_exp_NALM6_TC_vs_CC")
deg_tt_vs_cc <- run_deg(sf.df.data, sample_metadata, "CC", "TT", "DEG_gen_exp_NALM6_TT_vs_CC")

```

```{r}
library(pheatmap)

# 1. Extract top 10 upregulated genes by fpkm_TT
top_genes <- deg_tt_vs_cc %>%
    filter(significance == "down") %>%
    arrange(fpkm_TT) %>%
    slice_head(n = 25) %>%
    pull(gene_id)

# 2. Define NALM6 sample order: TT, TC, CC
nalm6_ordered <- c("NALM6_TT_1", "NALM6_TT_2", "NALM6_TT_3",
                   "NALM6_TC_1", "NALM6_TC_2", "NALM6_TC_3",
                   "NALM6_CC_1", "NALM6_CC_2", "NALM6_CC_3")

# 3. Subset FPKM matrix for selected genes and samples
heatmap_data <- fpkm_matrix_annotated %>%
    filter(gene_id %in% top_genes) %>%
    select(gene_name, all_of(nalm6_ordered)) %>%
    column_to_rownames("gene_name")

# 4. Log-transform for visualization
heatmap_matrix <- log2(heatmap_data + 1)

# 5. Plot heatmap without column clustering
pheatmap(heatmap_matrix,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         scale = "row",
         main = "Top 25 Downregulated Genes in TT vs CC (NALM6)",
         fontsize_row = 10)

```

## SPEARMAN CORRELATION of C dosage for NALM6


```{r}
# Add gene_id as a column
sf.df.data_symbols <- sf.df.data %>%
    tibble::rownames_to_column("gene_id")

# Join with tx2gene_df to get gene_name
sf.df.data_symbols <- sf.df.data_symbols %>%
    left_join(
        tx2gene_df %>%
            select(gene_id, gene_name) %>%
            distinct(gene_id, .keep_all = TRUE),
        by = "gene_id"
    ) %>%
    relocate(gene_name, .after = gene_id)



# Keep only NALM6 columns
nalm6_expr_matrix <- sf.df.data_symbols %>%
  select(gene_id, gene_name, starts_with("NALM6_"))

# 1. Extract counts matrix
nalm6_counts <- nalm6_expr_matrix %>%
  select(-gene_id, -gene_name) %>%
  as.matrix()
rownames(nalm6_counts) <- nalm6_expr_matrix$gene_id

# 2. Create sample metadata
nalm6_sample_metadata <- tibble(
  Sample = colnames(nalm6_counts)
) %>%
  mutate(original_sample = Sample) %>%  # preserve original name for joining
  separate(Sample, into = c("cell_line", "genotype", "replicate"), sep = "_") %>%
  mutate(genotype = factor(genotype, levels = c("TT", "TC", "CC")))

# 3. Create DESeq2 object
dds_nalm6 <- DESeqDataSetFromMatrix(
  countData = nalm6_counts,
  colData = nalm6_sample_metadata,
  design = ~1
)

# 4. Apply variance-stabilizing transformation
vst_nalm6 <- vst(dds_nalm6, blind = TRUE)

# 5. Extract transformed matrix
vst_matrix <- assay(vst_nalm6)

# 6. Add gene_id and gene_name
vst_df <- as.data.frame(vst_matrix) %>%
  tibble::rownames_to_column("gene_id") %>%
  left_join(nalm6_expr_matrix %>% select(gene_id, gene_name) %>% distinct(), by = "gene_id")

# 7. Reshape to long format and join metadata
vst_long <- vst_df %>%
  pivot_longer(cols = starts_with("NALM6_"), names_to = "Sample", values_to = "expression") %>%
  left_join(nalm6_sample_metadata, by = c("Sample" = "original_sample")) %>%
  mutate(dosage = case_when(
    genotype == "TT" ~ 0,
    genotype == "TC" ~ 1,
    genotype == "CC" ~ 2
  ))

# 🧮 Compute mean VST expression per genotype
mean_expr_per_genotype <- vst_long %>%
  group_by(gene_id, gene_name, genotype) %>%
  summarise(mean_expr = mean(expression), .groups = "drop") %>%
  pivot_wider(names_from = genotype, values_from = mean_expr,
              names_prefix = "mean_vst_expr_")

# 🔁 Compute Spearman correlation
dose_response_summary <- vst_long %>%
  group_by(gene_id, gene_name) %>%
  summarise(
    spearman_rho = cor(expression, dosage, method = "spearman"),
    .groups = "drop"
  ) %>%
  mutate(
    trend_direction = case_when(
      spearman_rho >= 0.5 ~ "↑ (increased with C dosage)",
      spearman_rho <= -0.5 ~ "↓ (decreased with C dosage)",
      TRUE ~ "– (weak/no trend)"
    )
  ) %>%
  left_join(mean_expr_per_genotype, by = c("gene_id", "gene_name")) %>%
  arrange(desc(abs(spearman_rho)))



```

### For REH Spearman

```{r}
# Add gene_id as a column
sf.df.data_symbols <- sf.df.data %>%
  tibble::rownames_to_column("gene_id")

# Join with tx2gene_df to get gene_name
sf.df.data_symbols <- sf.df.data_symbols %>%
  left_join(
    tx2gene_df %>%
      select(gene_id, gene_name) %>%
      distinct(gene_id, .keep_all = TRUE),
    by = "gene_id"
  ) %>%
  relocate(gene_name, .after = gene_id)

# 🎯 Subset only REH columns
reh_expr_matrix <- sf.df.data_symbols %>%
  select(gene_id, gene_name, starts_with("REH_"))

# 1. Extract counts matrix
reh_counts <- reh_expr_matrix %>%
  select(-gene_id, -gene_name) %>%
  as.matrix()
rownames(reh_counts) <- reh_expr_matrix$gene_id

# 2. Create sample metadata
reh_sample_metadata <- tibble(
  Sample = colnames(reh_counts)
) %>%
  mutate(original_sample = Sample) %>%
  separate(Sample, into = c("cell_line", "genotype", "replicate"), sep = "_") %>%
  mutate(genotype = factor(genotype, levels = c("TT", "TC", "CC")))

# 3. Create DESeq2 object
dds_reh <- DESeqDataSetFromMatrix(
  countData = reh_counts,
  colData = reh_sample_metadata,
  design = ~1
)

# 4. Apply variance-stabilizing transformation
vst_reh <- vst(dds_reh, blind = TRUE)

# 5. Extract transformed matrix
vst_matrix_reh <- assay(vst_reh)

# 6. Add gene_id and gene_name
vst_df_reh <- as.data.frame(vst_matrix_reh) %>%
  tibble::rownames_to_column("gene_id") %>%
  left_join(reh_expr_matrix %>% select(gene_id, gene_name) %>% distinct(), by = "gene_id")

# 7. Reshape to long format and join metadata
vst_long_reh <- vst_df_reh %>%
  pivot_longer(cols = starts_with("REH_"), names_to = "Sample", values_to = "expression") %>%
  left_join(reh_sample_metadata, by = c("Sample" = "original_sample")) %>%
  mutate(dosage = case_when(
    genotype == "TT" ~ 0,
    genotype == "TC" ~ 1,
    genotype == "CC" ~ 2
  ))

# 🧮 Compute mean VST expression per genotype
mean_expr_reh <- vst_long_reh %>%
  group_by(gene_id, gene_name, genotype) %>%
  summarise(mean_expr = mean(expression), .groups = "drop") %>%
  pivot_wider(names_from = genotype, values_from = mean_expr,
              names_prefix = "mean_vst_expr_")

# 🔁 Compute Spearman correlation
dose_response_summary_reh <- vst_long_reh %>%
  group_by(gene_id, gene_name) %>%
  summarise(
    spearman_rho = cor(expression, dosage, method = "spearman"),
    .groups = "drop"
  ) %>%
  mutate(
    trend_direction = case_when(
      spearman_rho >= 0.5 ~ "↑ (increased with C dosage)",
      spearman_rho <= -0.5 ~ "↓ (decreased with C dosage)",
      TRUE ~ "– (weak/no trend)"
    )
  ) %>%
  left_join(mean_expr_reh, by = c("gene_id", "gene_name")) %>%
  arrange(desc(abs(spearman_rho)))

# 🔍 Inspect results
dose_response_summary_reh %>% print(n = 10)

write_csv(dose_response_summary_reh,"/Users/cojulian/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/atacseq/Clarissas_bulk_ATAC_Aug_2025/both_rna_atac_reh_sig/reh_dose_response_summary_rnaseq.csv")

```

# Matt's requiest of Aug 18
First, identify TT vs CC DEGs using log2FC and p-value cutoffs; then, keep only those with meaningful FPKM in at least one genotype; finally, among those, retain genes where TC shows expression between TT and CC — indicating a graded, dosage-like effect.

## NALM6
```{r}
# Step 1: Keep genes with |log2FC| > 1 and pvalue < 0.05
de_genes_filtered <- de_results_tt_vs_cc %>%
  dplyr::filter(!is.na(pvalue)) %>%
  dplyr::filter(abs(log2FoldChange) > 1, pvalue < 0.05)

# Step 2: Extract and compute group means for TT and CC samples
fpkm_means <- fpkm_matrix_clean %>%
  dplyr::select(
    gene_id,
    starts_with("NALM6_TT_"),
    starts_with("NALM6_CC_")
  ) %>%
  rowwise() %>%
  mutate(
    mean_TT = mean(c_across(starts_with("NALM6_TT_")), na.rm = TRUE),
    mean_CC = mean(c_across(starts_with("NALM6_CC_")), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(gene_id, mean_TT, mean_CC)

# Step 3: Merge with DE gene list and apply FPKM filter
de_genes_with_means <- de_genes_filtered %>%
  dplyr::left_join(fpkm_means, by = "gene_id") %>%
  dplyr::filter(mean_TT > 1 | mean_CC > 1)

# Step 4: Add TC group mean
fpkm_means_tc <- fpkm_matrix_clean %>%
  dplyr::select(
    gene_id,
    starts_with("NALM6_TC_")
  ) %>%
  rowwise() %>%
  mutate(mean_TC = mean(c_across(starts_with("NALM6_TC_")), na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(gene_id, mean_TC)

# Join with previous table
final_genes <- de_genes_with_means %>%
  left_join(fpkm_means_tc, by = "gene_id") %>%
  # Step 5: Keep genes where TC falls between TT and CC
  dplyr::filter(
    (mean_TC > mean_TT & mean_TC < mean_CC) |
    (mean_TC < mean_TT & mean_TC > mean_CC)
  ) %>% 
  dplyr::select(-significance)

# Check how many genes passed all filters
cat("✅ Genes passing all filters:", nrow(final_genes), "\n")

readr::write_csv(final_genes, "/Users/cojulian/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/atacseq/Clarissas_bulk_ATAC_Aug_2025/both_rna_atac_nalm6_sig/filtered_DE_genes_TT_vs_CC_expression_screened.csv")
```

## REH

```{r}
# 📦 Load required libraries
library(DESeq2)
library(tidyverse)

# 1. Subset REH metadata for TT vs CC
meta_reh_tt_cc <- sample_metadata %>%
  filter(cell_line == "REH", genotype %in% c("TT", "CC")) %>%
  mutate(genotype = factor(genotype, levels = c("CC", "TT")))  # Reference = CC

# 2. Check and align sample IDs
reh_ids <- meta_reh_tt_cc$sample_id_full
missing_reh <- setdiff(reh_ids, colnames(sf.df.data))
if (length(missing_reh) > 0) warning("❗ Missing samples: ", paste(missing_reh, collapse = ", "))

# 3. Subset count matrix
counts_reh_tt_cc <- sf.df.data[, reh_ids, drop = FALSE]

# 4. Build DESeq2 object
dds_reh <- DESeqDataSetFromMatrix(
  countData = as.matrix(counts_reh_tt_cc),
  colData = meta_reh_tt_cc,
  design = ~ genotype
)

# 5. Run DESeq2
dds_reh <- DESeq(dds_reh)

# 6. Extract DE results (TT vs CC)
res_reh <- results(dds_reh, contrast = c("genotype", "TT", "CC"), alpha = 0.05)

# 7. Convert to dataframe and annotate
de_results_reh_tt_vs_cc <- as.data.frame(res_reh) %>%
  rownames_to_column("gene_id") %>%
  left_join(
    tx2gene_df %>% dplyr::select(gene_id, gene_name) %>% distinct(gene_id, .keep_all = TRUE),
    by = "gene_id"
  ) %>%
  mutate(
    FoldChange = 2^log2FoldChange,
    significance = case_when(
      FoldChange >= 1.5 & padj <= 0.05 ~ "up",
      FoldChange <= (1 / 1.5) & padj <= 0.05 ~ "down",
      TRUE ~ "ns"
    )
  )

# 8. Step 1: Filter DE genes by effect size and p-value
de_genes_filtered_reh <- de_results_reh_tt_vs_cc %>%
  filter(!is.na(pvalue)) %>%
  filter(abs(log2FoldChange) > 1, pvalue < 0.05)

# 9. Step 2: Compute mean expression (FPKM-like) for TT and CC
fpkm_means_reh <- fpkm_matrix_clean %>%
  select(gene_id, starts_with("REH_TT_"), starts_with("REH_CC_")) %>%
  rowwise() %>%
  mutate(
    mean_TT = mean(c_across(starts_with("REH_TT_")), na.rm = TRUE),
    mean_CC = mean(c_across(starts_with("REH_CC_")), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(gene_id, mean_TT, mean_CC)

# 10. Step 3: Merge with DE gene list and filter by mean expression
de_genes_with_means_reh <- de_genes_filtered_reh %>%
  left_join(fpkm_means_reh, by = "gene_id") %>%
  dplyr::filter(mean_TT > 1 | mean_CC > 1)

# 11. Step 4: Add mean TC expression
fpkm_means_tc_reh <- fpkm_matrix_clean %>%
  select(gene_id, starts_with("REH_TC_")) %>%
  rowwise() %>%
  mutate(mean_TC = mean(c_across(starts_with("REH_TC_")), na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(gene_id, mean_TC)

# 12. Step 5: Final filter (TC between TT and CC)
final_genes_reh <- de_genes_with_means_reh %>%
  left_join(fpkm_means_tc_reh, by = "gene_id") %>%
  filter((mean_TC > mean_TT & mean_TC < mean_CC) |
         (mean_TC < mean_TT & mean_TC > mean_CC)) %>%
  dplyr::select(-significance)

# ✅ Output summary
cat("✅ Genes passing all filters (REH TT vs CC):", nrow(final_genes_reh), "\n")

# 💾 Save to CSV
readr::write_csv(
  final_genes_reh,
  "/Users/cojulian/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/atacseq/Clarissas_bulk_ATAC_Aug_2025/both_rna_atac_reh_sig/filtered_DE_genes_REH_TT_vs_CC_expression_screened.csv"
)

```

