```{r}
# Load required libraries
suppressMessages(library(tidyverse))

# Define base directory containing sample folders
base_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/rnaseq/PolyA_Rnaseq_nickase/polyA_sf_files"

# Get full paths to sample directories (each folder = one clone)
sample_dirs <- list.dirs(base_dir, recursive = FALSE, full.names = TRUE)

# Extract sample folder names (e.g., NALM6_TT_Clone1)
sample_names <- basename(sample_dirs)

# Construct metadata dataframe
sample_metadata <- tibble(
    sample = sample_names,
    path   = file.path(sample_dirs, "quant.sf")
) %>%
    mutate(
        cell_line = "NALM6",   # all samples are NALM6
        genotype  = case_when(
            str_detect(sample, "TT") ~ "TT",
            str_detect(sample, "TC") ~ "TC",
            str_detect(sample, "CC") ~ "CC",
            TRUE ~ NA_character_
        ),
        clone     = str_extract(sample, "Clone[0-9]+") %>%
            str_remove("Clone") %>%
            as.integer(),
        replicate = 1,
        sample_id_full = paste(cell_line, genotype, ifelse(is.na(clone), "NA", clone), sep = "_")
    )

# Set rownames to match combined sample ID
rownames(sample_metadata) <- sample_metadata$sample_id_full

```


```{r}
suppressMessages(library(rtracklayer))

# Import GTF
gtf <- import("/Users/cojulian/Desktop/Tzu_projects/Matt/reference_genomces/Homo_sapiens.GRCh38.113.gtf.gz")
gtf_df <- as.data.frame(gtf)

# Build transcript-to-gene mapping
tx2gene_df <- gtf_df %>%
  select(transcript_id, gene_id, gene_name) %>%
  distinct(transcript_id, gene_id, gene_name) %>%
  drop_na()

```

## 🔹 Chunk 2. Align metadata for downstream analysis

```{r}
# Prepare colData for SummarizedExperiment / DESeq2
human_coldata <- sample_metadata %>%
  mutate(files = path) %>%
  dplyr::select(sample_id_full, files, everything()) %>%
  dplyr::rename(Sample = sample_id_full)

rownames(human_coldata) <- human_coldata$Sample

```


```{r}
# Initialize from first file
tmp <- read.delim(human_coldata$files[1]) %>%
  inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
  group_by(gene_id) %>%
  summarise(NumReads = as.integer(sum(NumReads)))

# Matrix dimensions
num.row <- nrow(tmp)
num.col <- nrow(human_coldata)

# Empty count matrix
sf.df <- data.frame(matrix(0, nrow = num.row, ncol = num.col))
colnames(sf.df) <- human_coldata$Sample
rownames(sf.df) <- tmp$gene_id

# Fill matrix
for (i in seq_along(human_coldata$files)) {
  cat("Reading:", human_coldata$files[i], "\n")
  
  sf.df[, i] <- read.delim(human_coldata$files[i]) %>%
    inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
    group_by(gene_id) %>%
    summarise(NumReads = as.integer(sum(NumReads))) %>%
    pull(NumReads)
}

```

```{r}
# Convert to matrix for downstream use (DESeq2, edgeR, etc.)
Salmon.matrix <- as.matrix(sf.df)

# Confirm sample IDs match
all(colnames(Salmon.matrix) %in% rownames(human_coldata))  # Should return TRUE
```
# CPM Filtering

CPM (counts per million) rescales raw counts by each library’s total reads, i.e., CPM = (raw_count / library_size) × 10^6, which directly adjusts for sequencing depth. edgeR’s filterByExpr() uses CPM thresholds that are adaptive to library sizes and your design (via group=), keeping genes that are sufficiently expressed in enough samples within groups to be testable in differential analysis. This approach is superior to a fixed raw-count cutoff because it avoids depth bias, reduces multiple-testing burden by removing uninformative, near-zero genes, and respects replication structure (e.g., TT/TC/CC triplicates). In short, CPM-aware filtering is principled, depth-aware, and aligned with best practice for RNA-seq DE pipelines.


```{r}
# ==============================================
# 🔹 CPM-aware filtering with edgeR::filterByExpr
# ==============================================

suppressMessages(library(edgeR))

# Build DGEList from Salmon.matrix
y <- DGEList(counts = Salmon.matrix)

# Define grouping (by genotype)
grp <- factor(
  human_coldata$genotype[match(colnames(Salmon.matrix), human_coldata$Sample)],
  levels = c("TT","TC","CC")
)

# Apply filterByExpr (CPM-aware, respects design and library size)
keep <- filterByExpr(y, group = grp)

cat("\n[filterByExpr summary]\n")
cat("Genes before filtering :", nrow(Salmon.matrix), "\n")
cat("Genes kept (CPM-aware) :", sum(keep), "\n")
cat("Percent retained        :", round(100 * sum(keep) / nrow(Salmon.matrix), 2), "%\n")

# Filter raw counts accordingly
sf.df.data <- Salmon.matrix[keep, ]

# ✅ Inspect result
cat("\nFinal CPM-filtered matrix dimensions:\n")
print(dim(sf.df.data))

cat("\nPreview filtered matrix:\n")
print(head(sf.df.data[, 1:5]))

```


```{r}
# ==============================================
# Per-sample distributions (pre vs CPM-filtered)
# ==============================================
suppressMessages(library(tidyverse))

# Long-form pre-filter (raw)
pre_long <- as.data.frame(Salmon.matrix) %>%
  tibble::rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "counts") %>%
  mutate(stage = "Unfiltered")

# Long-form post-filter (same raw counts, CPM-filtered gene set)
post_long <- as.data.frame(sf.df.data) %>%
  tibble::rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "counts") %>%
  mutate(stage = "CPM-filtered")

per_sample <- bind_rows(pre_long, post_long) %>%
  mutate(log_counts = log2(counts + 1)) %>%
  left_join(human_coldata %>% dplyr::select(Sample, genotype), by = "Sample")

ggplot(per_sample, aes(x = Sample, y = log_counts, fill = stage)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
  facet_grid(. ~ genotype, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Per-sample distributions (pre vs CPM-aware post-filter)",
    x = "Sample", y = "log2(counts + 1)", fill = "Stage"
  )

# ==============================================
# Post-filter logCPM (edgeR) per sample
# ==============================================
suppressMessages(library(edgeR))
suppressMessages(library(tidyverse))

# Compute logCPM on the CPM-filtered raw-count matrix
logcpm <- cpm(DGEList(counts = sf.df.data), log = TRUE, prior.count = 1)

logcpm_long <- as.data.frame(logcpm) %>%
  tibble::rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "logCPM") %>%
  left_join(human_coldata %>% dplyr::select(Sample, genotype), by = "Sample")

ggplot(logcpm_long, aes(x = Sample, y = logCPM)) +
  geom_boxplot(outlier.size = 0.3) +
  facet_grid(. ~ genotype, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Post-filter logCPM distributions per sample",
    x = "Sample", y = "logCPM"
  )


```


# Getting FPKM
```{r}
# ==============================================
# 🔹 Compute FPKM and plot ARID5B expression
# ==============================================

suppressMessages(library(DESeq2))
suppressMessages(library(rtracklayer))
suppressMessages(library(GenomicRanges))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))

# 1. Build DESeq2 object from filtered counts
dds <- DESeqDataSetFromMatrix(
  countData = round(sf.df.data),     # counts must be integers
  colData   = human_coldata,
  design    = ~ genotype
)
dds <- DESeq(dds)

# 2. Load GTF and compute exon union lengths
exons <- gtf[gtf$type == "exon"]

# Length per gene_id
exons_by_gene <- split(exons, exons$gene_id)
exonic_union  <- GenomicRanges::reduce(exons_by_gene)
gene_len_bp   <- vapply(exonic_union, function(gr) sum(width(gr)), integer(1))

# Attach basepairs to dds
rowData(dds)$basepairs <- gene_len_bp[rownames(dds)]

# 3. Compute FPKM
fpkm_mat <- fpkm(dds, robust = TRUE)
fpkm_df <- as.data.frame(fpkm_mat) %>%
  tibble::rownames_to_column("gene_id") %>%
  mutate(across(-gene_id, ~ round(.x, 3)))

# Preview
head(fpkm_df, 10)

# 4. Extract ARID5B (ENSG00000150347) expression
arid5b_expr <- fpkm_df %>%
  filter(gene_id == "ENSG00000150347") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "FPKM") %>%
  left_join(human_coldata, by = "Sample")   # no rownames_to_column needed

# 5. Plot expression across genotypes
ggplot(arid5b_expr, aes(x = genotype, y = FPKM, color = genotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.15, size = 3) +
  labs(title = "ARID5B Expression (FPKM)",
       x = "Genotype", y = "FPKM (log scale)") +
  scale_y_log10() +
  theme_minimal()

```

# 🔹 PCA for NALM6

```{r}
# ==============================================
# PCA for NALM6 samples
# ==============================================
suppressMessages(library(tidyverse))
suppressMessages(library(tidybulk))
suppressMessages(library(ggrepel))

# 1. Filter metadata for NALM6 samples
nalm6_coldata <- human_coldata %>%
  filter(cell_line == "NALM6")

# 2. Filter expression matrix to match NALM6 samples
nalm6_expr <- sf.df.data[, nalm6_coldata$Sample, drop = FALSE] %>%
  tibble::rownames_to_column("gene_id")

# 3. Reshape to tidy format
nalm6_expr_tidy <- nalm6_expr %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "Sample",
    values_to = "abundance"
  ) %>%
  rename(gene = gene_id)

# 4. Merge with metadata
nalm6_tidy <- nalm6_expr_tidy %>%
  left_join(nalm6_coldata, by = "Sample")

# 5. Tidybulk PCA
pca_nalm6 <- nalm6_tidy %>%
  tidybulk(
    .sample = Sample,
    .transcript = gene,
    .abundance = abundance
  ) %>%
  keep_abundant() %>%
  scale_abundance() %>%
  reduce_dimensions(method = "PCA", .dims = 8)

# 6. PCA plot: color by genotype, shape by clone
pca_nalm6 %>%
  pivot_sample() %>%
  ggplot(aes(
    x = PC1, y = PC2,
    color = genotype,
    shape = as.factor(clone),
    label = Sample
  )) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(max.overlaps = 5, size = 3, show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PCA of NALM6 Gene Expression",
    subtitle = "Colored by Genotype, Shaped by Clone",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )

```

# ✅ Updated QC Suite (SummarizedExperiment + plots)

```{r}
# ==============================================
# Build SummarizedExperiment with filtered counts
# ==============================================
suppressMessages(library(SummarizedExperiment))

# Ensure colData matches sample order
sf.df.se <- SummarizedExperiment(
  assays = list(counts = as.matrix(sf.df.data)),
  colData = human_coldata[colnames(sf.df.data), ]
)

# Add detailed sample names
sf.df.se$detailed_name <- paste(
  sf.df.se$cell_line,
  sf.df.se$genotype,
  sf.df.se$clone,
  sep = "_"
)

# ==============================================
# Library size plot
# ==============================================
library_sizes <- colSums(assay(sf.df.se))

ggplot(
  data.frame(Sample = sf.df.se$detailed_name, LibrarySize = library_sizes),
  aes(x = Sample, y = LibrarySize)
) +
  geom_bar(stat = "identity", fill = "skyblue") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold")
  ) +
  labs(
    title = "Library Size Distribution",
    x = "Sample (cell_line_genotype_clone)",
    y = "Library Size (Raw Counts)"
  )

# ==============================================
# Log2 expression distributions per sample
# ==============================================
suppressMessages(library(reshape2))

# Log-transform
log_counts <- log2(assay(sf.df.se) + 1)

# Melt to long format
log_counts_long <- reshape2::melt(log_counts)
colnames(log_counts_long) <- c("Gene", "Sample", "LogExpression")

# Annotate by detailed name
log_counts_long <- log_counts_long %>%
  mutate(Group = sf.df.se$detailed_name[match(Sample, colnames(log_counts))])

# Density plot by sample
ggplot(log_counts_long, aes(x = LogExpression, color = Group)) +
  geom_density(size = 1, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Log2 Gene Expression Distribution by Sample",
    x = "Log2(Counts + 1)",
    y = "Density"
  ) +
  theme(
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9)
  )


```
# # Doing DE, only one comparison

```{r}
# ==============================================
# DESeq2 Analysis: NALM6 TT vs CC
# ==============================================
suppressMessages(library(DESeq2))
suppressMessages(library(dplyr))
suppressMessages(library(tibble))

# 1. Filter metadata to only NALM6 TT and CC clones
meta_data <- sample_metadata %>%
  dplyr::filter(cell_line == "NALM6", genotype %in% c("TT", "CC")) %>%
  dplyr::mutate(genotype = factor(genotype, levels = c("CC", "TT")))  # CC = reference

# 2. Subset count matrix
filtered_bulk_matrix <- sf.df.data[, meta_data$sample_id_full]

# 3. Build DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(filtered_bulk_matrix),
  colData   = meta_data,
  design    = ~ genotype
)

# 4. Run DESeq2
dds <- DESeq(dds)

# 5. Extract results: TT vs CC
res <- results(dds, contrast = c("genotype", "TT", "CC"), alpha = 0.05)

# 6. Tidy results
de_results_tt_vs_cc <- as.data.frame(res) %>%
  tibble::rownames_to_column("gene_id") %>%
  left_join(
    tx2gene_df %>%
      select(gene_id, gene_name) %>%
      distinct(gene_id, .keep_all = TRUE),
    by = "gene_id"
  ) %>%
  mutate(FoldChange = 2^log2FoldChange) %>%
  mutate(significance = case_when(
    FoldChange >= 1.5 & padj <= 0.05 ~ "up",
    FoldChange <= (1 / 1.5) & padj <= 0.05 ~ "down",
    TRUE ~ "ns"
  ))

# 7. Summarize DE results
cat("\n[DESeq2 TT vs CC summary]\n")
summary(res)

```

```{r}
# ==============================================
# Volcano plot: TT vs CC
# ==============================================
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))

volcano_data <- de_results_tt_vs_cc %>%
  mutate(GeneSymbol = ifelse(is.na(gene_name), gene_id, gene_name)) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  select(GeneSymbol, log2FoldChange, padj, FoldChange, significance) %>%
  mutate(GeneType = significance)

# Plot styling
cols   <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey80")
sizes  <- c("up" = 2.5, "down" = 2.5, "ns" = 1.2)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.6)

ggplot(volcano_data, aes(
  x = log2(FoldChange),
  y = -log10(padj),
  fill = GeneType,
  size = GeneType,
  alpha = GeneType
)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(log2(1 / 1.5), log2(1.5)), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  ggrepel::geom_text_repel(
    data = filter(volcano_data, GeneType != "ns"),
    aes(label = GeneSymbol),
    size = 4,
    color = "darkblue",
    max.overlaps = 10,
    min.segment.length = Inf
  ) +
  labs(
    title = "Volcano Plot: NALM6 TT vs CC (padj ≤ 0.05, FC ≥ 1.5)",
    x = "log2(Fold Change)",
    y = "-log10(Adjusted P-value)"
  ) +
  theme_minimal(base_family = "Arial") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12)
  )

```

## Venn diagram of polyA and ribo0
```{r}
# ==============================================
# Venn diagrams: PolyA vs Ribo-0 DEG comparison
# ==============================================
suppressMessages(library(tidyverse))
suppressMessages(library(ggvenn))
suppressMessages(library(cowplot))

# 1. Load external DEG results (Ribo-0)
external_res <- read_csv(
  "~/Desktop/Tzu_projects/Matt/Clarissa/snp_cell_lines_public_datasets_project/cell_lines/rnaseq/Gene_expresion_analysis/deg_results_NALM6/DEG_gen_exp_NALM6_TT_vs_CC.csv"
)

# 2. Define UP/DOWN sets from current DESeq2 results (PolyA)
polyA_up <- de_results_tt_vs_cc %>%
  filter(significance == "up") %>%
  pull(gene_id)

polyA_down <- de_results_tt_vs_cc %>%
  filter(significance == "down") %>%
  pull(gene_id)

# 3. Define UP/DOWN sets from external results (Ribo-0)
ribo0_up <- external_res %>%
  filter(significance == "up") %>%
  pull(gene_id)

ribo0_down <- external_res %>%
  filter(significance == "down") %>%
  pull(gene_id)

# 4. Build lists for ggvenn
venn_up_list <- list(PolyA = polyA_up, Ribo0 = ribo0_up)
venn_down_list <- list(PolyA = polyA_down, Ribo0 = ribo0_down)

# 5. Create UP Venn plot
venn_up_plot <- ggvenn(
  venn_up_list,
  fill_color = c("#ffad73", "#26b3ff"),
  stroke_size = 1.5,
  set_name_size = 8,
  text_size = 8
) +
  ggtitle("Upregulated Genes: PolyA vs Ribo-0") +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5)
  )

# 6. Create DOWN Venn plot
venn_down_plot <- ggvenn(
  venn_down_list,
  fill_color = c("#ffad73", "#26b3ff"),
  stroke_size = 1.5,
  set_name_size = 8,
  text_size = 8
) +
  ggtitle("Downregulated Genes: PolyA vs Ribo-0") +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5)
  )

# 7. Combine plots vertically (no parentheses in labels)
combined_plot <- plot_grid(
  venn_up_plot, venn_down_plot,
  label_size = 24,
  label_fontface = "bold",
  ncol = 1,
  align = "v"
)

# 8. Display combined plot
print(combined_plot)

# Find overlapping genes
overlap_up <- intersect(polyA_up, ribo0_up)
overlap_down <- intersect(polyA_down, ribo0_down)

# Create data frame with direction
overlap_df <- tibble(
    gene_id = c(overlap_up, overlap_down),
    direction = c(rep("up", length(overlap_up)), rep("down", length(overlap_down)))
)

# Join with GTF to get gene symbols
overlap_annotated <- overlap_df %>%
    left_join(
        gtf_df %>% select(gene_id, gene_name) %>% distinct(),
        by = "gene_id"
    ) %>%
    select(gene_id, gene_name, direction)
# Save to CSV
write_csv(overlap_annotated, "overlapping_genes_both_RNA_technologies.csv")


```


